---
title: "Reproducibility & project workflow in Python"
subtitle: "ZAS Pythonistas working group"
date: today
date-format: "MMMM YYYY"
format:
  docx:
    toc: false
    number-sections: false
    highlight-style: github
  pdf:
    toc: false
    number-sections: false
    highlight-style: github
    geometry: "top=2cm, bottom=2cm, left=2cm, right=2cm"
    fontsize: 10pt
    papersize: a4
    keep-tex: false
    pdf-engine: xelatex
    mainfont: "Helvetica Neue"
    sansfont: "Helvetica Neue"
    monofont: "Menlo"
    colorlinks: true
    linkcolor: linkblue
    urlcolor: linkblue
    include-in-header:
      text: |
        \usepackage{float}
        \floatplacement{table}{H}
        \usepackage{longtable}
        \usepackage{xcolor}
        \usepackage{colortbl}
        \usepackage{tcolorbox}
        \tcbuselibrary{skins, breakable}
        \definecolor{linkblue}{HTML}{4472C4}
        \definecolor{codebg}{HTML}{F3F3F3}
        \definecolor{codeborder}{HTML}{BBBBBB}
        \definecolor{labelterm}{HTML}{3C3C3C}
        \definecolor{labelpython}{HTML}{3572A5}
        \definecolor{labelyaml}{HTML}{CB171E}
        \definecolor{labelmd}{HTML}{083FA1}
        \definecolor{stripebg}{HTML}{F2F6FC}
        \definecolor{headerbg}{HTML}{4472C4}
        \definecolor{highlightbg}{HTML}{D9E8F7}
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| output: false
library(knitr)
is_pdf <- knitr::is_latex_output()

kt <- function(df,
               col_names     = colnames(df),
               col_widths    = NULL,
               bold_col      = NULL,
               red_col       = NULL,
               green_col     = NULL,
               highlight     = NULL,
               highlight_col = "#D9E8F7") {

  if (is_pdf) {

    n_cols <- ncol(df)

    # Fallback: equal column widths summing to 16.5cm (A4 minus 2cm margins each side)
    if (is.null(col_widths))
      col_widths <- rep(paste0(round(16.5 / n_cols, 2), "cm"), n_cols)

    # p{} columns force text wrapping and avoid tabu/longtabu entirely
    col_spec <- paste0("p{", col_widths, "}", collapse = "")

    # Escape special LaTeX characters FIRST so backslashes in
    # \textbf{} and \textcolor{} injected below are not double-escaped
    df <- as.data.frame(lapply(df, function(x) {
      x <- as.character(x)
      x <- gsub("_",  "\\\\_",  x)   # underscore -- math subscript trigger
      x <- gsub("&",  "\\\\&",  x)   # ampersand
      x <- gsub("%",  "\\\\%",  x)   # percent
      x <- gsub("#",  "\\\\#",  x)   # hash
      x
    }))

    # Wrap cell values in LaTeX colour/bold commands AFTER escaping
    wrap_col <- function(d, idx, cmd) {
      d[[idx]] <- paste0(cmd, "{", d[[idx]], "}")
      d
    }
    if (!is.null(bold_col))  df <- wrap_col(df, bold_col,  "\\textbf")
    if (!is.null(red_col))   df <- wrap_col(df, red_col,   "\\textcolor[HTML]{993C1D}")
    if (!is.null(green_col)) df <- wrap_col(df, green_col, "\\textcolor[HTML]{2E7D52}")

    # Build raw LaTeX via kable — plain tabular, no longtable, no tabu
    raw <- as.character(kable(df,
      col.names = col_names,
      format    = "latex",
      booktabs  = FALSE,
      linesep   = "",
      longtable = FALSE,
      escape    = FALSE
    ))

    # Swap kable's auto column spec for our explicit p{} widths
    raw <- gsub(
      "\\\\begin\\{tabular\\}\\{[^}]+\\}",
      paste0("\\\\setlength\\{\\\\tabcolsep\\}\\{4pt\\}\n\\\\begin\\{tabular\\}\\{", col_spec, "\\}"),
      raw
    )

    # Colour the header row: inject \rowcolor + \color after the first \hline
    raw <- sub(
      "(\\\\hline\n)([^\n]+\\\\\\\\\n)(\\\\hline)",
      paste0("\\1\\\\rowcolor[HTML]{4472C4}\\\\color{white}\\\\bfseries\\\\small \\2\\3"),
      raw
    )

    # Alternating stripe: walk body rows and colour every second one
    lines      <- strsplit(raw, "\n")[[1]]
    hline_pos  <- grep("^\\\\hline$", lines)
    header_end <- if (length(hline_pos) >= 2) hline_pos[2] else NA

    if (!is.na(header_end)) {
      body      <- seq(header_end + 1, length(lines))
      row_lines <- grep("\\\\\\\\", lines[body]) + header_end
      even_rows <- row_lines[seq(2, length(row_lines), 2)]
      lines[even_rows] <- paste0("\\rowcolor[HTML]{F2F6FC}", lines[even_rows])

      if (!is.null(highlight) && highlight <= length(row_lines)) {
        hl <- row_lines[highlight]
        col_hex <- gsub("#", "", highlight_col)
        lines[hl] <- sub(
          "^(\\\\rowcolor[^}]*\\})?",
          paste0("\\\\rowcolor[HTML]{", col_hex, "}"),
          lines[hl]
        )
      }
      raw <- paste(lines, collapse = "\n")
    }

    # Wrap in a [H] float with small font
    raw <- paste0("\\begin{table}[H]\n\\small\n", raw, "\n\\end{table}")

    structure(raw, class = "knit_asis", knit_meta = list())

  } else {
    # docx: plain kable only
    kable(df, col.names = col_names)
  }
}
```

## Session overview

**Duration:** 90 minutes. **Format:** Discussion + live demo + hands-on exercise.

This session builds from general principles to Python-specific tools -- concepts that apply in any language first, then the specific tools and file formats you will use day-to-day in Python.

---

## Schedule

```{r}
#| label: tbl-schedule
schedule <- data.frame(
  Section = c(
    "1. What is reproducibility?",
    "2. Virtual environments & file paths",
    "3. Naming conventions & folder structure",
    "4. Version control with git",
    "5. Quarto vs. Jupyter notebooks",
    "6. Hands-on exercise",
    "7. Wrap-up & next topic vote"
  ),
  Scope = c(
    "Language-agnostic",
    "Python-specific",
    "Language-agnostic",
    "Language-agnostic",
    "Python-specific",
    "Python",
    "--"
  ),
  Time = c("8 min", "18 min", "12 min", "10 min", "17 min", "18 min", "7 min")
)
kt(schedule,
   col_widths = c("9cm", "5cm", "2cm"),
   highlight  = 6)
```

---

## 1 - What is reproducibility?

Reproducibility means someone else -- or future you -- can re-run your analysis and get the same result. It operates at two levels:

- **Environment reproducibility:** the same packages and versions are installed. Your code does not break because a collaborator has a different version of pandas.
- **Workflow reproducibility:** the same code runs in the same order and produces the same outputs. No mystery variables, no deleted raw files, no steps that exist only in someone's memory.

Neither layer is sufficient alone. A perfectly locked environment does not help if the script requires you to run cell 7 before cell 3. Clean, ordered code does not help if it requires a package version nobody can install anymore.

**Common failure modes:** "it works on my machine" *(environment)*; raw data was overwritten and the analysis cannot be re-run *(workflow)*; a notebook gives different results depending on the order cells were run *(workflow)*; a script uses an absolute path like `/Users/name/projects/...` that breaks on every other machine *(workflow)*.

Reproducibility is a spectrum -- the goal is not perfection from day one, but deliberate choices that reduce friction for your future self and collaborators.

---

## 2 - Virtual environments & file paths

### Virtual environments

A virtual environment is an isolated Python installation scoped to a single project -- the R equivalent of an `renv` library. Do not install packages into your global Python.

**This working group uses conda.** venv + pip is shown as a lighter alternative for pure-Python projects.

```{r}
#| label: tbl-conda
conda_tbl <- data.frame(
  Question = c(
    "Create an environment",
    "Install packages",
    "Record env for sharing",
    "Restore environment",
    "Lockfile (approx. renv.lock)",
    "When to use"
  ),
  conda = c(
    "conda create -n my-env python=3.11",
    "conda install pandas  or  pip install pandas",
    "conda env export --from-history > environment.yml",
    "conda env create -f environment.yml",
    "environment.yml or conda-lock for full pinning",
    "Non-Python deps, HPC clusters, mixed R+Python"
  ),
  venv_pip = c(
    "python -m venv .venv",
    "pip install pandas",
    "pip freeze > requirements.txt",
    "pip install -r requirements.txt",
    "requirements.txt or pyproject.toml + uv.lock",
    "Pure-Python projects, lighter setup"
  )
)
kt(conda_tbl,
   col_names  = c("", "conda (current setup)", "venv + pip (alternative)"),
   col_widths = c("3.5cm", "6.5cm", "6.5cm"),
   bold_col   = 1)
```

> **`--from-history` vs full export:** `conda env export` pins every package including transitive dependencies and platform-specific build strings, making it non-portable across operating systems. `--from-history` records only what you explicitly installed, letting conda resolve the rest per platform. For sharing within a team, `--from-history` is usually the better choice.

**A note on terminals: Mac vs Windows**

All code in this handout is written for **bash** (the shell used on macOS and Linux). Windows uses PowerShell by default, which has different syntax. Windows users should use one of the following so the commands here work as written:

```{r}
#| label: tbl-terminals
#| output: asis
terminals <- data.frame(
  Option = c(
    "Git Bash",
    "WSL (Windows Subsystem for Linux)",
    "Positron / VS Code integrated terminal",
    "Anaconda Prompt"
  ),
  Description = c(
    "Installs with Git for Windows. Gives a full bash shell. Simplest option for this session.",
    "A full Linux environment inside Windows. More powerful but requires one-time setup.",
    "Auto-uses bash or Git Bash if configured. Conda commands work out of the box.",
    "A Windows terminal pre-configured for conda. Conda commands work but bash syntax may not."
  ),
  Recommended = c("Yes -- simplest", "Optional", "Yes", "Conda only")
)
kt(terminals,
   col_names  = c("Option", "Description", "For this session"),
   col_widths = c("3.5cm", "9.5cm", "3cm"))
```

**Basic conda workflow:**

```{.bash filename="Terminal (bash / Git Bash / WSL)"}
conda env list                                  # list all environments
conda create -n zas-python-wg python=3.11       # create once
conda activate zas-python-wg                    # activate
conda info --envs                               # check active environment (asterisk = active)
conda install pandas matplotlib                 # install packages
conda env export --from-history > environment.yml  # record environment
conda env create -f environment.yml             # restore (others)
```

> **Positron / VS Code:** Command Palette -> *Python: Select Interpreter* -> choose your conda environment. The integrated terminal will use your active conda environment automatically.

### IDEs: VS Code and Positron

Both VS Code and Positron are built on the same engine (VS Code's core), so the interface is nearly identical. Positron is Posit's fork tuned for data science: it adds a persistent Variables pane, a Console panel, and a Plots viewer -- all familiar from RStudio. For this working group either IDE works; Positron is recommended if you are coming from RStudio.

```{r}
#| label: tbl-ide
ide_tbl <- data.frame(
  Feature = c(
    "Built on",
    "Python interpreter selection",
    "Environment activation",
    "Variables pane",
    "Console panel",
    "Plots viewer",
    "Notebook support",
    "R support",
    "Recommended for"
  ),
  VSCode = c(
    "VS Code core",
    "Command Palette -> Python: Select Interpreter",
    "Auto-activates .venv or conda env in workspace",
    "Via Python extension (limited)",
    "Integrated terminal only",
    "Via Jupyter extension",
    "Full .ipynb support",
    "Via R extension",
    "General development, large teams"
  ),
  Positron = c(
    "VS Code core (Posit fork)",
    "Same -- Command Palette -> Python: Select Interpreter",
    "Same -- auto-activates on project open",
    "Built-in, always visible (like RStudio)",
    "Built-in, persistent (like RStudio)",
    "Built-in viewer pane",
    "Full .ipynb support",
    "First-class, built-in",
    "Data science, R users moving to Python"
  )
)
kt(ide_tbl,
   col_names  = c("", "VS Code", "Positron"),
   col_widths = c("3.5cm", "6.5cm", "6cm"),
   bold_col   = 1)
```

```{r}
#| label: workflow-tip
#| output: asis
workflow_pdf <- "\\begin{tcolorbox}[colback=highlightbg, colframe=linkblue, arc=3pt, title={\\sffamily\\bfseries Your typical workflow}, fonttitle=\\bfseries\\small\\sffamily, coltitle=white, boxed title style={colback=linkblue}, breakable]
\\small
\\begin{enumerate}
  \\item Open Positron (or VS Code)
  \\item \\textbf{File $\\to$ Open Folder} $\\to$ navigate to your project folder
  \\item Check the interpreter in the \\textbf{bottom status bar} -- it should show your conda environment name (e.g. \\texttt{zas-python-wg}). If not: Command Palette $\\to$ Python: Select Interpreter $\\to$ choose it
  \\item Open the integrated terminal (\\texttt{Ctrl+`} or Terminal $\\to$ New Terminal) -- it should activate your conda environment automatically
  \\item Verify: \\texttt{conda info --envs} (asterisk next to active env) and \\texttt{pwd} (correct folder)
\\end{enumerate}
\\end{tcolorbox}"

workflow_docx <- "> **Your typical workflow**
>
> 1. Open Positron (or VS Code)
> 2. **File -> Open Folder** -> navigate to your project folder
> 3. Check the interpreter in the **bottom status bar** -- it should show your conda environment name (e.g. `zas-python-wg`). If not: Command Palette -> Python: Select Interpreter -> choose it
> 4. Open the integrated terminal (`Ctrl+backtick` or Terminal -> New Terminal) -- it should activate your conda environment automatically
> 5. Verify with `conda info --envs` (asterisk = active env) and `pwd` (correct folder)"

if (is_pdf) cat(workflow_pdf) else cat(workflow_docx)
```

```{r}
#| label: uv-callout
#| output: asis
if (is_pdf) {
  cat("\\begin{tcolorbox}[colback=highlightbg, colframe=linkblue, arc=3pt,
    title={\\sffamily\\bfseries uv -- a faster modern alternative worth knowing about},
    fonttitle=\\bfseries\\small\\sffamily, coltitle=white,
    boxed title style={colback=linkblue}, breakable]
\\small
\\texttt{uv} is a package and project manager for Python written in Rust (by the team behind \\texttt{ruff}).
It replaces venv, pip, pip-tools, and pyenv with a single tool and is 10-100x faster than pip for installs.

\\medskip
\\textbf{What makes it different from venv + pip:}
\\begin{itemize}
  \\item Generates a proper lockfile (\\texttt{uv.lock}) with full dependency graph resolution -- a much closer analogue to \\texttt{renv.lock} than \\texttt{pip freeze} is
  \\item Manages Python versions itself: \\texttt{uv python install 3.12} -- no separate pyenv needed
  \\item Scaffolds new projects: \\texttt{uv init my-project}
  \\item Runs scripts without manual activation: \\texttt{uv run script.py}
\\end{itemize}

\\medskip
\\textbf{Basic workflow (Terminal):}
\\begin{verbatim}
uv init my-project        # scaffold project + pyproject.toml
cd my-project
uv add pandas matplotlib  # install + update lockfile automatically
uv sync                   # restore environment from uv.lock
\\end{verbatim}

See: \\url{https://docs.astral.sh/uv}
\\end{tcolorbox}")
} else {
  cat("> **uv -- a faster modern alternative worth knowing about**
>
> `uv` is a package and project manager for Python written in Rust (by the team behind `ruff`).
> It replaces venv, pip, pip-tools, and pyenv with a single tool and is 10-100x faster than pip.
>
> Key differences from venv + pip: generates a proper lockfile (`uv.lock`, closer to `renv.lock` than `pip freeze`);
> manages Python versions itself (`uv python install 3.12`); scaffolds projects (`uv init my-project`);
> runs scripts without activation (`uv run script.py`).
>
> Basic workflow: `uv init my-project` / `uv add pandas matplotlib` / `uv sync`
>
> See: <https://docs.astral.sh/uv>")
}
```

### Project-relative file paths

Hardcoded absolute paths break on every machine except the one they were written on. There are two good ways to fix this in Python.

**Option 1: `pyprojroot` (recommended -- closest to `here::here()`)**

Install once: `conda install pyprojroot` or `pip install pyprojroot`

```{.python filename="Python"}
from pyprojroot import here
from pathlib import Path

data_path   = here("data/raw/stimuli.csv")
output_path = here("outputs/figures")
```

`pyprojroot` walks up from the current file until it finds a project root marker -- `.git`, `pyproject.toml`, `environment.yml`, etc. -- and anchors all paths there. This is the direct Python equivalent of `here::here()` and works correctly regardless of where the script lives in the project.

**Option 2: `pathlib` (no extra install)**

```{.python filename="Python"}
from pathlib import Path

# __file__ is the path of the current script
# .parent navigates up one folder at a time -- count carefully
PROJECT_ROOT = Path(__file__).parent.parent

data_path   = PROJECT_ROOT / "data" / "raw" / "stimuli.csv"
output_path = PROJECT_ROOT / "outputs" / "figures"
output_path.mkdir(parents=True, exist_ok=True)  # create if needed
```

This works without any extra packages but requires counting `.parent` calls correctly -- which breaks if you move the script to a different subfolder depth. Use `pyprojroot` if you want `here()`-style convenience.

---

## 3 - Naming conventions & folder structure

These conventions apply in any language. Consistent structure makes projects navigable by collaborators and comprehensible to your future self.

### Recommended folder layout

```{.bash filename="Project structure"}
my-project/
  data/
    raw/          # original data -- never modified
    processed/    # derived data, reproducible from raw
  scripts/        # numbered analysis scripts
  notebooks/      # exploratory .ipynb files
  outputs/
    figures/
    tables/
  docs/           # manuscripts, protocols, notes
  environment.yml # conda environment (or requirements.txt)
  .gitignore
  README.md
```

**Raw data is sacred.** Never overwrite it, never modify it in place. All transformations happen in code and produce outputs in `data/processed/`.

### File naming

```{r}
#| label: tbl-naming
naming <- data.frame(
  Rule = c(
    "Lowercase only",
    "No spaces",
    "Date-prefix versioned outputs",
    "Verb-first for scripts",
    "Zero-pad numbered scripts",
    "Be specific"
  ),
  Avoid = c(
    "My Analysis.py",
    "model results final.csv",
    "results.csv",
    "data.py",
    "1_clean.py, 10_model.py",
    "analysis.py, data2.csv"
  ),
  Prefer = c(
    "my-analysis.py",
    "model-results-final.csv",
    "2025-05-04_model-results.csv",
    "01_clean-data.py",
    "01_clean.py, 10_model.py",
    "01_clean-eyetracking.py"
  )
)
kt(naming,
   col_widths = c("4cm", "6cm", "6cm"),
   red_col    = 2,
   green_col  = 3)
```

### README as a map

Every project should answer three questions: what does this project do, how do I run it, and what do the outputs mean?

```{.markdown filename="README.md"}
# Project name
One sentence describing the study or task.

## Setup
conda env create -f environment.yml
conda activate zas-pythonistas

## Running the analysis
python scripts/01_clean-data.py
python scripts/02_fit-model.py

## Outputs
outputs/figures/  -- all plots, referenced in the manuscript
outputs/tables/   -- model results tables
```

---

## 4 - Version control with git

Git tracks changes to your files over time, so you can see what changed, when, and why -- and roll back if something breaks. It is language-agnostic and works for any text-based project. GitHub (and GitLab, Codeberg, etc.) are hosting platforms for git repositories; git itself runs entirely on your local machine.

### Core concepts

- **Repository (repo):** a folder tracked by git, containing your project and its full history
- **Commit:** a saved snapshot of your project at a point in time, with a message describing what changed
- **Staging area:** a holding zone where you choose exactly which changes to include in the next commit -- you do not have to commit everything at once
- **Branch:** a parallel version of the project where you can develop a feature or fix without affecting the main version (more on this below)

### The basic workflow

```{.bash filename="Terminal"}
# --- One-time setup ---
git config --global user.name  "Your Name"
git config --global user.email "you@example.com"

# --- Start tracking a project ---
git init                        # initialise a new repo in the current folder

# --- Day-to-day workflow ---
git status                      # see what has changed
git add scripts/01_analyse.py   # stage a specific file
git add .                       # stage everything (use carefully)
git commit -m "fix: use relative paths in analysis script"

# --- Review history ---
git log --oneline               # compact list of past commits
git diff                        # see unstaged changes line by line
```

### Writing useful commit messages

A good commit message completes the sentence "If applied, this commit will...". Use a short imperative verb phrase: `add`, `fix`, `update`, `remove`. Keep it under 72 characters.

```{r}
#| label: tbl-commits
commits <- data.frame(
  Avoid = c(
    "fixed stuff",
    "update",
    "changes",
    "asdfgh"
  ),
  Prefer = c(
    "fix: use relative paths in analysis script",
    "add: conda environment file and README setup instructions",
    "remove: hardcoded absolute paths from all scripts",
    "refactor: split data cleaning into separate script"
  )
)
kt(commits,
   col_widths = c("4cm", "12.5cm"),
   red_col    = 1,
   green_col  = 2)
```

### What to put in .gitignore

```{.bash filename=".gitignore"}
# Python
.venv/
__pycache__/
*.pyc
*.pyo
.ipynb_checkpoints/

# Conda
.conda/

# Secrets -- never commit these
.env
*.key

# Large data files (store elsewhere, e.g. OSF or NAS)
data/raw/
*.csv
*.edf
*.mat
```

### A note on branching

Branches let you work on a new feature or experiment without touching the main version of your project. When the work is ready, you merge it back. For solo projects, you may not need branches at all -- a clean linear commit history is fine. For collaborative projects, a simple convention is: keep `main` stable and do new work on a named branch.

```{.bash filename="Terminal"}
git branch feature/add-preprocessing   # create a branch
git checkout feature/add-preprocessing # switch to it
                                        # (or: git checkout -b name, to do both at once)
git checkout main                       # switch back to main
git merge feature/add-preprocessing    # merge when ready
```

We will cover branching and collaborative workflows (pull requests, merge conflicts) in a future session.

---

## 5 - Quarto vs. Jupyter notebooks

### They are not competitors

Quarto *uses* Jupyter as its execution engine -- adding `jupyter: python3` to a `.qmd` header tells Quarto to call Jupyter to run your code. The real choice is between two **file formats**:

```{r}
#| label: tbl-formats
formats <- data.frame(
  Dimension = c(
    "File format",
    "Execution order",
    "Kernel state",
    "Git diffs",
    "Output storage",
    "Multi-format export",
    "Inline computed values",
    "Best for"
  ),
  Notebook = c(
    ".ipynb (JSON)",
    "Any order -- cells run independently",
    "Persists between cells; can be stale",
    "Noisy -- outputs embedded in JSON",
    "Stored inside the file",
    "Requires nbconvert + manual steps",
    "Not straightforward",
    "Exploration, iteration, teaching"
  ),
  Quarto = c(
    ".qmd (plain text)",
    "Always top-to-bottom, fresh process",
    "No persistent state between renders",
    "Clean -- plain text only",
    "Separated from source",
    "Built-in: HTML, Word, PDF, slides",
    "`{python} round(val, 2)`",
    "Reports, papers, shared outputs"
  )
)
kt(formats,
   col_names  = c("", ".ipynb notebook", ".qmd (Quarto)"),
   col_widths = c("3.5cm", "6.5cm", "6.5cm"),
   bold_col   = 1)
```

### The hidden state problem in notebooks

The biggest reproducibility risk in `.ipynb` files is **out-of-order execution**. Jupyter remembers every variable set during a session, even from cells you have deleted or moved. A notebook that appears to work may depend on a variable set hours ago in a cell that no longer exists. The only reliable check is *Kernel -> Restart & Run All* before sharing -- and people often forget.

### What `freeze: true` does in Quarto

By default, Quarto re-runs all your code from scratch on every render -- correct for reproducibility, but slow for heavy computations like model fitting or EEG preprocessing. `freeze: true` changes this: Quarto **caches each chunk's output** and only re-runs a chunk when its source code actually changes.

- **First render:** runs everything, saves outputs into a `_freeze/` folder alongside your `.qmd`
- **Subsequent renders:** skips unchanged chunks and reuses saved outputs
- **After editing a chunk:** only that chunk re-runs

Committing `_freeze/` to git means collaborators can render the document and see your outputs without needing your data or compute environment.

```{.yaml filename="_quarto.yml"}
execute:
  freeze: true
```

> **R parallel:** `freeze: true` is roughly analogous to `knitr cache = TRUE` in R Markdown, but more reliable -- it caches at the file level, tracks source changes explicitly, and plays well with git.

### A practical two-stage workflow

Use notebooks for the messy exploratory work. Once the approach is settled, move clean code into a `.qmd` for the shareable output.

```{.bash filename="Terminal"}
# Extract clean code from a notebook when you are ready
jupyter nbconvert --to script exploration.ipynb
# produces exploration.py -- paste the useful parts into your .qmd
```

---

## 6 - Hands-on exercise

**Setup:** the group repo is already cloned and your conda environment is active. The repo contains `scripts/01_analyse.py` (a short script with a reproducibility problem) and `environment.yml`.

### Part A - Dependency management

```{r}
#| label: tbl-exercise-a
ex_a <- data.frame(
  Step = c("A1", "A2", "A3", "A4"),
  Task = c(
    "Activate your conda environment: conda activate zas-pythonistas",
    "Export it: conda env export > environment-full.yml -- open the file and inspect what is there",
    "Compare with the leaner version: conda env export --from-history > environment-minimal.yml -- what is the difference?",
    "Optional: try the venv equivalent in a separate terminal: python -m venv .venv && source .venv/bin/activate && pip install pandas && pip freeze > requirements.txt"
  )
)
kt(ex_a, col_widths = c("1.2cm", "15.3cm"))
```

### Part B - Project-relative file paths

Open `scripts/01_analyse.py`. It contains a hardcoded absolute path that breaks on every machine except the one it was written on.

```{r}
#| label: tbl-exercise-b
ex_b <- data.frame(
  Step = c("B1", "B2", "B3"),
  Task = c(
    "Run the script and read the error: python scripts/01_analyse.py",
    "Fix the broken path using pathlib so it resolves relative to the project root (see the pathlib pattern on this handout)",
    "Run the script again to confirm it works: python scripts/01_analyse.py"
  )
)
kt(ex_b, col_widths = c("1.2cm", "15.3cm"))
```

### Part C - Git hygiene

```{r}
#| label: tbl-exercise-c
ex_c <- data.frame(
  Step = c("C1", "C2", "C3", "Stretch"),
  Task = c(
    "Check what git would currently commit: git status -- notice anything that should not be there?",
    "Create a .gitignore that excludes: __pycache__/, *.pyc, .ipynb_checkpoints/, .venv/ (if created in A4)",
    "Stage your fixed script and .gitignore and commit with a descriptive message",
    "Move the output of 01_analyse.py into a .qmd file and render it to HTML with Quarto"
  )
)
kt(ex_c,
   col_widths = c("1.5cm", "15cm"),
   highlight  = 4)
```

### Skeleton script for the repo

Add this as `scripts/01_analyse.py` before the session. Also commit a small `data/raw/sample.csv` with a few rows of dummy data.

```{.python filename="scripts/01_analyse.py"}
import pandas as pd
from pathlib import Path

# -------------------------------------------------------
# TODO: this path is broken -- fix it using pathlib
# -------------------------------------------------------
data_path = "/Users/replace-me/projects/zas-pythonistas/data/raw/sample.csv"

df = pd.read_csv(data_path)
print(f"Loaded {len(df)} rows")
print(df.describe())
```

---

## R to Python parallels

```{r}
#| label: tbl-reference
ref <- data.frame(
  Concept = c(
    "IDE",
    "Isolated environment",
    "Package install",
    "Dependency lockfile",
    "Restore environment",
    "Project root",
    "Project-relative paths",
    "Set random seed",
    "Dynamic report format",
    "Freeze / cache computations",
    "Inline computed value",
    "Version control"
  ),
  R = c(
    "RStudio / Positron",
    "renv library (.Rprofile)",
    "install.packages() / pak::pak()",
    "renv.lock",
    "renv::restore()",
    ".Rproj file sets working directory",
    "here::here()",
    "set.seed(42)",
    ".Rmd or .qmd",
    "knitr cache = TRUE",
    "`r round(val, 2)`",
    "git (same)"
  ),
  Python = c(
    "Positron / VS Code",
    "conda env  or  .venv (venv)",
    "conda install / pip install",
    "environment.yml  or  requirements.txt  or  uv.lock",
    "conda env create -f environment.yml  or  pip install -r requirements.txt",
    "No automatic root -- use explicit Path(__file__) convention",
    "here(\"data/raw/file.csv\") via pyprojroot  or  Path(__file__).parent / ...",
    "random.seed(42)  +  numpy.random.seed(42)",
    ".qmd with jupyter: python3",
    "freeze: true in Quarto",
    "`{python} round(val, 2)`",
    "git (same)"
  )
)
kt(ref, col_widths = c("3.8cm", "4.8cm", "7.9cm"))
```

---

## Resources

- **pythRon book** (ZAS internal): R/Python concepts side-by-side with worked examples
- **Quarto + Python:** [quarto.org/docs/computations/python.html](https://quarto.org/docs/computations/python.html)
- **Quarto freeze:** [quarto.org/docs/projects/code-execution.html#freeze](https://quarto.org/docs/projects/code-execution.html#freeze)
- **uv:** [docs.astral.sh/uv](https://docs.astral.sh/uv)
- **Pro Git book (free):** [git-scm.com/book/en/v2](https://git-scm.com/book/en/v2)
- **Real Python -- virtual environments primer:** [realpython.com/python-virtual-environments-a-primer](https://realpython.com/python-virtual-environments-a-primer/)
- **The Turing Way** -- community handbook for reproducible research: [the-turing-way.netlify.app](https://the-turing-way.netlify.app)
