---
title: "Debug me"
engine: jupyter
knitr: true
---

## Task {.unnumbered}

Run this script. Several things will break. Find and fi each error, one at a time.

::: {.callout-tip}
Read the error messages carefully -- python usually tells you exaction what is wrong and on which line.
:::

## Imports

::: {.callout-note}
### python version of `here`

```{python}
from pyprojroot import here
```

:::

```{python}
# python                 # R equivalent
import numpy as np       # baseR
import pandas as pd      # tidyverse; mainly dplyr and tibble
import seaborn as sns    # ggplot2
```

```{r}
#| code-fold: true
#| code-summary: "R equivalent"
# install.packages("pacman")
pacman::p_load(dplyr, tibble, ggplot2, readr, here)
```

## Load data

Make sure you are running the script from the project root, or adjust the path below.

```{python}
#| eval: false
df = pd.read_csv("data/scores.csv")
```

```{r}
#| code-fold: true
#| code-summary: "R equivalent"
df <- read_csv(here("data","scores.csv"))
```

::: {.callout-note}
### My fix

I don't have a working directory set. Do that.

```{python}
import os
os.getcwd()
```

Wrong, that wasn't the problem. The problem was that data isn't in the project parent node; it's nested within sessions/march27; but i can also load it in using here from pyprojroot which uses wherever a .git file is as the project root

```{python}
df = pd.read_csv(here("data/scores.csv"))
```

OR, change the wd to be the `sessions/march-27` folder. The equivalent to `setwd()` in python is:

```{python}
#| eval: false
os.chdir("sessions/march-27")
```

I will *not* use this option though. My preferred project structure would have `data/` in the project folder head.


:::

## Quick look

```{python}
print("Shape:", df.shape)
print(df.head())
```

```{r}
#| code-fold: true
#| code-summary: "R equivalent"
length(df)
nrow(df)
head(df)
```

## Filter rows

```{python}
#| eval: false
# Keep only rows where condition is TRUE
good_rows <- df["score"] > 50        

print("Rows passing filter:", len(good_rows))
```

My fix: change `<-` (R syntax) to `=` (also works in R).

```{python}
# Keep only rows where condition is TRUE
good_rows = df[df["score"] > 50]    

print("Rows passing filter:", len(good_rows))
```


```{r}
#| code-fold: true
#| code-summary: "R equivalent"
good_rows <-
    df |>
    filter(score > 50)

paste("Rows passing filter:", nrow(good_rows))
```

## Compute a summary

```{python}
mean_score = df["score"].mean()
max_score  = df["score"].max()

print("Mean score:", round(mean_score, 2)) 
print("Max score: ", max_score)
```

```{r}
#| code-fold: true
#| code-summary: "R equivalent"
mean_score <- mean(df$score)
max_score <- max(df$score)

paste("Mean score:", round(mean_score,2))
paste("Max score:", max_score)
```

## Add a new column

Fix: wrap whole expression in brackets and remove forward slash.

```{python}
# Normalise scores to 0–1 range
df["score_norm"] = (
    (df["score"] - df["score"].min()) /
    (df["score"].max() - df["score"].min()) 
)

```

```{r}
#| code-fold: true
#| code-summary: "R equivalent"

df$score_norm = (df$score - min(df$score)) / (max(df$score) - min(df$score))
```

# Plot

```{python}
sns.histplot(data = df, x = "score_norm")
```

```{r}
#| code-fold: true
#| code-summary: "R equivalent"

hist(df$score_norm)
```
