---
title: "Translate R to Python"
subtitle: "ZAS Python Working Group — Session 1, Exercise 2"
format: html
engine: jupyter
knitr: true
---

# Assignment operator

```{r}
#| eval: false
participant_count <- 12
session_num <- 1
```

```{python}
participant_count = 12
session_num = 1
```

# Logical

All caps in R.

```{r}
#| eval: false
is_recorded <- TRUE
has_video   <- FALSE
```

Capitalised in Python.

```{python}
is_recorded = True
has_video   = False
```

# Indexing

R: concatenate with `c()` and indexing begins at 1. Last element with `length(x)`.
```{r}
#| eval: false
scores <- c(72, 85, 61, 90, 78)
first_score <- scores[1]
last_score  <- scores[length(scores)]
```

Python: concatenate with `[]` and indexing begins at 1. To extract last element use `-1`.
```{python}
scores = [72, 85, 61, 90, 78]
first_score = scores[0]
last_score  = scores[-1]
```

# Paste

In R: `cat()` removes numerical prefix (`[1]`) added by `paste()`. New line with `\n`.

```{r}
#| eval: false
cat(paste("Participant count:", participant_count, "\n"))
cat(paste("Session:", session_num, "\n"))
cat(paste("Mean score:", mean(scores), "\n"))
```

Python: `print()` adds new line automatically. f-string handles concatenataion. Using `scores.mean()` won't work here because `scores` is a plain Python list, not a numpy array or pandas series (which it would be if above it was e.g., `scores = np.array([72, 85, 61, 90, 78])`).

```{python}
print(f"Participant count: {participant_count}")
print(f"Session: {session_num}")
print(f"Participant count: {sum(scores) / len(scores)}")
```

# Function

```{r}
#| eval: false
normalise <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}
```

Python: `def` to define a function with explictly state what to `return`
```{python}
def normalise(x):
    return (x - min(x)) / (max(x) - min(x))
```

```{r}
#| eval: false
normalised_scores <- normalise(scores)
cat(paste("Normalised:", normalised_scores, "\n"))
```

Python: we need to convert to a numpy array first because Python lists don't support arthimetic operations. This is a huge difference between Python and R. Normally we'd import `numpy` at the top of the script, but we'll keep it here for now.

```{python}
import numpy as np
scores = np.array([72, 85, 61, 90, 78])
normalised_scores = normalise(scores)
```

:::{.callout-tip}
## Python is general-purpose

R was designed from the ground up for statistical computing, so statistical and mathematical operations are baked into the language. Python is a general-purpose programming language not specifically intended for data analysis. Statistical functionality is added through libraries like `numpy`, `pandas`, and `scipy`. That's why Python is ideal for data processing and R is ideal for analyses.

:::
